% Acknwoledgements page
\thispagestyle{empty}

\begin{center}
    \Large
    
\section*{Acknowledgements}

\end{center}

\vspace{1cm}

First, I would like to thank my thesis advisor Prof.\ Duane Nykamp for  his guidance, invaluable advise and constructive criticism while  writing this paper and throughout my academic journey. The door to Prof.\ Nykamp's office was always open whenever I ran into a trouble spot or had a question about my research or writing. He  allowed this paper to be my own work, but steered me in the right the direction whenever he thought I needed it.


My very special thanks to  the experts who were involved in the examination and evaluation for this research project: Prof.\ Richard McGehee and Prof.\ Galin Jones.  Without their passionate participation and input,  my plan A master's thesis could not have been successfully completed.


I would  like to acknowledge Madeline Handschy as the second reader of this thesis, and I am gratefully indebted to her for her very valuable comments on this thesis.


Finally, I must express my very profound gratitude to my beloved husband, Dr.  Norbert Tambe Ebai  for providing me with unfailing love and continuous encouragement throughout my years in graduate school, and my precious children Kathleen and Neilton, who are the pride and joy of my life. This accomplishment would not have been possible without them.
Thanks to my family and friends for encouraging me to follow my dreams. Special thanks to my father, John Mark Otim, who encouraged me to pursue farther studies and who has been by my side through thick and thin.
Last but not least, words cannot express my deep gratitude for my Guardians Laveeda and Lynn Battle for their
unwaivering support and for encouraging me to believe in my self.






\vfill




\newpage